Working document to try to explain why/how ATAs influence coexistence in an intuitive way for referee 3

### 2023-07-28 paper/ExplanationOfATAEffectsWorkingDoc.tex
ongoing work on a set of intuitive explanations and examples for ATA effects

\subsection{Dan thoughts 2023 07 26}

@@

%Attempt 2 below - better than the above, but may need to be simplified a bit more,
%if possible
%
Just as for storage effects, temporal niche differentiation is necessary for ATA
effects. This makes intuitive sense because temporal niche differentiation is 
also a prerequistite of storage effects [\cite{Chesson_1994} and Box 2 of \cite{StumpVasseur2023}],
and ATA effects are a part of storage effects, but see
SI section X for a detailed explanation. 
%
Assuming that temporal niche differentiation is present, how do ATA effects then happen?
We here offer an intuitive explanation. 
%
If temporal niche differentiation occurs
but is partial, the $E$ and $C$ experienced by a currently scarce 
species (i.e., $E_i$ and $C_{i \bs i})$ will be positively but imperfectly associated, 
and the association may be asymmetric. 
%
Storage effects can occur if some buffering mechanism reduces losses in bad years so that
gains in good years can more than make up for them. 
%
Bad years happen when the negative effects of $C$ outweigh any 
positive effects of $E$; but this can occur
both 1) when $E$ is moderately favorable but $C$ is very strong, as well as 2) when $E$ is very unfavorable,
or moderately unfavorable and $C$ is not too weak.
Conversely, good years happen when positive effects of $E$ outweigh negative effects of $C$;
but this can occur 1) when $E$ is very favorable, or moderately favorable and $C$ is not too strong,
as well as 2) when $E$ is moderately unfavorable but $C$ is very weak. 
%
It turns out that the potential for storage to occur when 
$E$ and $C$ are both higher than average (the cases 1, above); versus the potential  
when $E$ and $C$ are
both lower than average (the cases 2) are not necessarily equivalent.
Depending on the growth rate function, $r(E,C)$, storage potential may differ
across the spectrum from 1 to 2.
%
ATA effects are the extent to which asymmetries of association between $E$ and 
$C$ are aligned with the potential for storage across the dichotomy 1 versus 2.
A comparison is made to a benchmark of symmetric association between $E$ and $C$.
%
ATA effects therefore help quantify different degrees of temporal niche differentiation 
occuring under different sets of environmental conditions (such as cases 1 and 2), possibly via
different ecological or physiological mechanisms.
%
Details, applications, and examples of this intuition 
are in SI section X, and these ideas are also used there to provide 
intuitive explanations for the relative differences in the importance
of ATA effects between our log-normal and beta fecundities lottery models.
%DAN: You ought to also mention explaining differences with the diatom system,
%if you figure out an explanation.

\subsection{SI section X corresponding to the thoughts above}

Just as for storage effects [see \cite{Chesson_1994} and Box 2 of \cite{StumpVasseur2023}], 
temporal niche differentiation is necessary in order for ATA
effects to materialize; we here explain why in detail. ``Lack of temporal niche differentiation'' can be interpreted
to mean that species respond in identical ways to environmental fluctuations. 
In a two-species model such as those we consider, this should mean a monotonically
increasing relationship between $E_i$ and $C_{i \bs i}$. The former is the environment
as perceived by species $i$. The latter is competition that a scarce invader $i$ experiences
in a monoculture of $j$, and that competition comes only from $j$ (because $i$ itself
is scarce), so should correspond to the environmental response of $j$. If 
$E_i$ and $C_{i \bs i}$ are monotonically related, then $(E_i,C_{i \bs i})$
and $(E_i^\nat,C_{i \bs i}^\nat)$ are distributed in the same way, so 
$\E[r_i(E_i,c_{i \bs i})]-\E[r_i(E_i^\nat,c_{i \bs i}^\nat)]=0$.
Therefore, 
\begin{align}
\Delta_i^{[EC]} &= (\olr_{i \bs i}-\olr_{j \bs i})-(\olr_{i \bs i}^\nat - \olr_{j \bs i}^\nat) \\
&= \olr_{i \bs i}-(\olr_{i \bs i}^\nat - \olr_{j \bs i}^\nat) \\
&= \E[r_i(E_i,C_{i \bs i})]-\E[r_i(E_i^\nat,C_{i \bs i}^\nat)]+\E[r_j(E_j^\nat,C_{j \bs i}^\nat)] \label{eq:expri}\\
&= \E[r_j(E_j^\nat,C_{j \bs i}^\nat)].
\end{align}
But we already know $\E[r_j(E_j^\nat,C_{j \bs i}^\nat)]=0$
because there is no niche differentiation between $j$ and itself, so 
$E_i$ and $C_{j \bs i}$ should be monotonically related. So 
$\E[r_j(E_j^\nat,C_{j \bs i}^\nat)] = \E[r_j(E_j,C_{j \bs i})]=0$, giving
$\Delta_i^{[EC]}=0$. So temporal
niche differentiation is necessary for ATA effects on coexistence to 
materialize.

If temporal niche differentiation occurs but is partial, 
then $E_i$ and $C_{i \bs i}$ will be positively but imperfectly correlated with
each other, corresponding to similar but non-identical responses of species to
environmental fluctuation; but how do ATA effects then happen? We here extend the intuition
introduced in the Discussion and add detail.

Storage can occur if population
increases during good years outweigh decreases during bad years. For instance,
if $r_i(E_i,C_{i \bs i})$ is very large when the balance of $E_i$ and 
$C_{i \bs i}$ is favorable, but only moderately small when this balance is
unfavorable, the expected value $\E[r_i(E_i,C_{i \bs i})]$ in (\ref{eq:expri})
will be positive. But the potential for such storage of population growth
may differ between when $E_i$ and $C_{i \bs i}$ are both high compared
to when they are both low. 
For instance, for the beta fecundities lottery model, $E_i/C_{i \bs i}$, a quantity
which features prominently in $\E[r_i(E_i,C_{i \bs i})]=\log(1-\delta+E_i/C_{i \bs i})$,
takes values bounded above by $\frac{2\eta_i \delta}{\eta_j}$ 
and bounded below by $\frac{\eta_i \delta}{2 \eta_j}$ when $E_i$ and $C_{i \bs i}$
are both in the upper halves of their assumed beta distribtions (recall that 
these quantities were assumed to be distributed, respectively, as $\eta_1 \beta$
and $\eta_2 \beta/\delta$ for $\beta$ a beta-distributed random variable with parameters 
$0.5$ and $0.5$). These bounds are separated only by a factor of four. 
On the other hand, when $E_i$ and $C_{i \bs i}$ are both in the lower halves 
of their distributions, $E_i/C_{i \bs i}$ can take any positive value because
then either $E_i$ or $C_{i \bs i}$ can take values arbitrarily close to $0$. 
Because values of $E_i/C_{i \bs i}$ have the potential to be much larger when 
$E_i$ and $C_{i \bs i}$ are both in the lower halves of their distributions
than when these quantities are both in their upper halves, and because
the growth rate, $\log(1-\delta+E_i/C_{i \bs i})$, buffers low values of 
$E_i/C_{i \bs i}$ (the infemum of the growth rate is $\log(1-\delta)$, whereas the supremum is
infinite), the storage effects contributed by the left tails of 
$E_i$ and $C_{i \bs i}$ have the potential to be much greater than those contributed
by the right tails. Essentially, values of $\log(1-\delta+E_i/C_{i \bs i})$
occurring when $E_i$ and $C_{i \bs i}$ are both in the upper halves of their
distributions can never exceed $\log(1-\delta+\frac{2\eta_i \delta}{\eta_j})$,
whereas values ocurring when $E_i$ and $C_{i \bs i}$ are both in the lower
halves of their distributions can be arbitrarily large (though with decreasing 
probability for larger values).

>>>>

\subsection{Dan thoughts 2023 07 26 - planned for inclusion in the Discussion}

@@

%Attempt 2 below - better than the above, but may need to be simplified a bit more,
%if possible
%
Just as for storage effects, temporal niche differentiation is necessary for ATA
effects. This makes intuitive sense because temporal niche differentiation is 
also a prerequistite of storage effects [\cite{Chesson_1994} and Box 2 of \cite{StumpVasseur2023}],
and ATA effects are a part of storage effects, but see
SI section X1 for a detailed explanation. 
%
Assuming that temporal niche differentiation is present, how do ATA effects then happen?
We here offer an intuitive explanation. 
%
If temporal niche differentiation occurs
but is partial, the $E$ and $C$ experienced by a currently scarce 
species (i.e., $E_i$ and $C_{i \bs i})$ will be positively but imperfectly associated, 
and the association may be asymmetric. 
%
Storage effects can occur if some buffering mechanism reduces losses in bad years so that
gains in good years can more than make up for them. 
%
Bad years happen when the negative effects of $C$ outweigh any 
positive effects of $E$; but this can occur
both 1) when $E$ is moderately favorable but $C$ is very strong, as well as 2) when $E$ is very unfavorable,
or moderately unfavorable and $C$ is not too weak.
Conversely, good years happen when positive effects of $E$ outweigh negative effects of $C$;
but this can occur 1) when $E$ is very favorable, or moderately favorable and $C$ is not too strong,
as well as 2) when $E$ is moderately unfavorable but $C$ is very weak. 
%
It turns out that the potential for storage to occur when 
$E$ and $C$ are both higher than average (the cases 1, above); versus the potential  
when $E$ and $C$ are
both lower than average (the cases 2) are not necessarily equivalent.
Depending on the growth rate function, $r(E,C)$, storage potential may differ
across the spectrum from 1 to 2.
%
ATA effects are the extent to which asymmetries of association between $E$ and 
$C$ are aligned with the potential for storage across the dichotomy 1 versus 2.
If $E$ and $C$ are strongly associated in the parts of their distribution where
storage potential is high, ATA effects will be relatively low, and vice versa.
A comparison is made to a benchmark of symmetric association between $E$ and $C$.
%
ATA effects therefore help quantify different degrees of temporal niche differentiation 
occuring under different sets of environmental conditions (such as cases 1 and 2), possibly via
different ecological or physiological mechanisms.
%
Details, applications, and examples of this intuition 
are in SI section X2, and these ideas are also used there to provide 
intuitive explanations for the relative differences in the importance
of ATA effects between our log-normal and beta fecundities lottery models.
%DAN: You ought to also mention explaining differences with the diatom system,
%if you figure out an explanation.

\subsection{SI section X1 corresponding to the thoughts above, call it Temporal niche differentiation is necessary for ATAs}\label{SIsect:temporal_niche_diff}

Just as for storage effects [see \cite{Chesson_1994} and Box 2 of \cite{StumpVasseur2023}], 
temporal niche differentiation is necessary in order for ATA
effects to materialize; we here explain why, in detail. ``Lack of temporal niche differentiation'' can be interpreted
to mean that species respond in identical ways to environmental fluctuations. 
In a two-species model such as those we consider, this should mean a monotonically
increasing relationship between $E_i$ and $C_{i \bs i}$. The former is the environment
as perceived by species $i$. The latter is competition that a scarce invader $i$ experiences
in a monoculture of $j$, and that competition comes only from $j$ (because $i$ itself
is scarce), so should correspond to the environmental response of $j$. If 
$E_i$ and $C_{i \bs i}$ are monotonically related, then $(E_i,C_{i \bs i})$
and $(E_i^\nat,C_{i \bs i}^\nat)$ are distributed in the same way, so 
$\E[r_i(E_i,c_{i \bs i})]-\E[r_i(E_i^\nat,c_{i \bs i}^\nat)]=0$.
Therefore, 
\begin{align}
\Delta_i^{[EC]} &= (\olr_{i \bs i}-\olr_{j \bs i})-(\olr_{i \bs i}^\nat - \olr_{j \bs i}^\nat) \\
&= \olr_{i \bs i}-(\olr_{i \bs i}^\nat - \olr_{j \bs i}^\nat) \\
&= \E[r_i(E_i,C_{i \bs i})]-\E[r_i(E_i^\nat,C_{i \bs i}^\nat)]+\E[r_j(E_j^\nat,C_{j \bs i}^\nat)] \label{eq:expri}\\
&= \E[r_j(E_j^\nat,C_{j \bs i}^\nat)].
\end{align}
But we already know $\E[r_j(E_j^\nat,C_{j \bs i}^\nat)]=0$
because there is no niche differentiation between $j$ and itself, so 
$E_j$ and $C_{j \bs i}$ should be monotonically related. So 
$\E[r_j(E_j^\nat,C_{j \bs i}^\nat)] = \E[r_j(E_j,C_{j \bs i})]=0$, giving
$\Delta_i^{[EC]}=0$. So temporal
niche differentiation, interpreted as Spearman correlation between 
$E$ and $C$ less than $1$, is necessary for ATA effects on coexistence to 
materialize.

\subsection{SI section X2 corresponding to the thoughts above, call it Intuition and 
examples for ATAs}\label{SIsect:intuition_examples}

If temporal niche differentiation occurs but is partial, 
then $E_i$ and $C_{i \bs i}$ will be positively but imperfectly correlated with
each other, corresponding to similar but non-identical responses of species to
environmental fluctuation (see SI section \ref{SIsect:temporal_niche_diff}); 
but how do ATA effects then happen? We here extend the intuition
introduced in the Discussion and provide examples.

Storage can occur if population
increases during good years outweigh decreases during bad years. For instance,
if $r_i(E_i,C_{i \bs i})$ is very large when the balance of $E_i$ and 
$C_{i \bs i}$ is favorable, but only moderately small when this balance is
unfavorable, the expected value $\E[r_i(E_i,C_{i \bs i})]$ in (\ref{eq:expri})
will be positive. But the potential for such storage of population growth
may differ between when $E_i$ and $C_{i \bs i}$ are both high compared
to when they are both low. 
For instance, for the beta fecundities lottery model, $E_i/C_{i \bs i}$, a quantity
which features prominently in $r_i(E_i,C_{i \bs i})=\log(1-\delta+E_i/C_{i \bs i})$,
takes values bounded above by $\frac{2\eta_i \delta}{\eta_j}$ 
and bounded below by $\frac{\eta_i \delta}{2 \eta_j}$ when $E_i$ and $C_{i \bs i}$
are both in the upper halves of their assumed beta distribtions (recall that 
these quantities were assumed to be distributed, respectively, as $\eta_1 \beta$
and $\eta_2 \beta/\delta$ for $\beta$ a beta-distributed random variable with parameters 
$0.5$ and $0.5$). These bounds are separated only by a factor of four. 
On the other hand, when $E_i$ and $C_{i \bs i}$ are both in the lower halves 
of their distributions, $E_i/C_{i \bs i}$ can take any positive value because
then either $E_i$ or $C_{i \bs i}$ can take values arbitrarily close to $0$. 
Because values of $E_i/C_{i \bs i}$ have the potential to be much larger when 
$E_i$ and $C_{i \bs i}$ are both in the lower halves of their distributions
than when these quantities are both in their upper halves, and because
the growth rate, $\log(1-\delta+E_i/C_{i \bs i})$, buffers low values of 
$E_i/C_{i \bs i}$ (the infemum of the growth rate is $\log(1-\delta)$, whereas the supremum is
infinite), the storage effects contributed by the left tails of 
$E_i$ and $C_{i \bs i}$ have the potential to be much greater than those contributed
by the right tails. Essentially, values of $\log(1-\delta+E_i/C_{i \bs i})$
occurring when $E_i$ and $C_{i \bs i}$ are both in the upper halves of their
distributions can never exceed $\log(1-\delta+\frac{2\eta_i \delta}{\eta_j})$,
whereas values ocurring when $E_i$ and $C_{i \bs i}$ are both in the lower
halves of their distributions can be arbitrarily large (though with decreasing 
probability for larger values). Apparently, the years with the most potential to 
contribute to storage
effects for the beta-fecundities lottery model are those for which the environment
is actually poorer than average, because, for those years, competition can be very low,
so that the ratio $E/C$ is large. Years for which the
environment is exceptionally good actually cannot contribute much to storage effects for 
this model, because for those years, $E/C$ cannot be large.
It is the imbalance between $E$ and $C$ that allows for storage effects. 
Counterintuitively, for this model, that imbalance is greatest when the environment is 
poor. This result is counterintuitive because one may be used to thinking of storage effects
occuring when reproductive pulses from environmentally good years sustain the population
through environmentally poor years. But, in fact, good years for population growth may
happen principally when the environment is poor, if those are also the times when 
competition is very low, as for this model.

Fig. \ref{fig:beta_fecundidities_explanatory} provides quantitative backing to
the ideas of the prior paragraph. The growth function $\ln(1-\delta+E/C)$ that pertains 
to the lottery model
is plotted with colors and contours on Fig. \ref{fig:beta_fecundidities_explanatory}A-C.
Samples were taken from the distribution $(E_i,C_{i \bs i})$ for the beta 
fecundities lottery model for left- and right-tailed 
associations between $E_i$ and $C_{i \bs i}$, and are plotted as
points on Fig. \ref{fig:beta_fecundidities_explanatory}A and C, respectively. 
Samples were also taken from the distribution $(E_i^\nat,C_{i \bs i}^\nat)$
(which is the same regardless of whether the original $(E_i,C_{i \bs i})$
distribution showed left- or right-tail association) and are plotted on 
Fig. \ref{fig:beta_fecundidities_explanatory}B. 
The dashed lines labelled $L_2$ on A-C separate the upper-right $50\%$
of each distribution from the lower-left $50\%$. The dashed lines labeled
$L_1$ separate the upper-left $50\%$ from the lower-right $50\%$.
The quantity $\E \left[ \ln \left( 1-\delta+\frac{E_i}{C_{i \bs i}} \right) \right]$ can be
well approximated for left- and right-tail associations by averaging function 
values, depicted with color and contours, over the points on panels A and C, respectively. The quantity
$\E \left[ \ln \left( 1-\delta+\frac{E_i^{||}}{C_{i \bs i}^{||}} \right) \right]$
can be well approximated by averaging function 
values over the points on B. So ATA effects for the model with left-tail association
between $E_i$ and $C_{i \bs i}$ (respectively, right-tail association) are given by 
computing the average function value over points on panel A (respectively, C)
and then subtracting the average function value over points on panel B.
Contributions to the storage effect of points in the left portions of 
the $E,C$ distribution (the region labeled ``left'' on panels A and C)
can be accurately estimated by averaging function values across
points in that region; likewise storage contributions of points in the 
``right'' region can be estimated by averaging function values over 
points in that region.

The details of the averaging process described above are pictured in 
Fig. \ref{fig:beta_fecundidities_explanatory}, panels D-F;
we now describe how. 
Signed distances of points on Fig. \ref{fig:beta_fecundidities_explanatory} 
A-C from the lines $L_1$ correspond to 
the quantities $\log_{10}(E_1/C_{1 \bs 1})$ on A, C and to 
$\log_{10}(E_1^\nat/C_{1 \bs 1}^\nat)$ on B. The distributions of these 
quantities across points to the left of $L_2$ are pictured 
in the lower horizontal-axis historgrams on panels D-F, respectively;
and distributions for points to the right of $L_2$ are 
pictured in the upper horizontal-axis histograms. 
For the left-tailed case (A, D), distributions to the

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/PostHocExplanatoryFig_betaFecund.jpg}
\caption{A detailed look at ATA effects for the beta fecundities lottery model.
The figure gives a pictorial representation of the calculation of  
$\E \left[ \ln \left( 1-\delta+\frac{E_i}{C_{i \bs i}} \right) \right]$ (A, C, D, F) and 
$\E \left[ \ln \left( 1-\delta+\frac{E_i^{||}}{C_{i \bs i}^{||}} \right) \right]$ (B, E)
for the model, considering both left- (A, D) and right-tail (C, F) associations
between $E_i$ and $C_{i \bs i}$.
The colors and contours on A-C $\ln(1-\delta+E/C)$, whereas the points show a sample
from the distribution $(E_i,C_{i \bs i})$ (A, C) or $(E_i^{||},C_{i \bs i}^{||})$ (B).
The quantity $\E \left[ \ln \left( 1-\delta+\frac{E_i}{C_{i \bs i}} \right) \right]$ can be
well approximated for left- and right-tail associations by averaging function 
values over the points on panels A and C, respectively. The quantity
$\E \left[ \ln \left( 1-\delta+\frac{E_i^{||}}{C_{i \bs i}^{||}} \right) \right]$
can be well approximated by averaging function 
values over the points on B. Values of $\log_{10}(E/C)$ for the points on A-C correspond 
to signed distances of the points from the upward sloping dashed lines. 
We used parameters $\eta_1 = 1$, $\eta_2 = 1.2$, and and $\delta=0.6$. }\label{fig:beta_fecundidities_explanatory}
\end{figure}
